---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "man/figures/README-",
  out.width  = "100%",
  out.height = "100%",
  message    = FALSE,
  warning    = FALSE
)

options(width = 160)
```

# AnnotateCells

[![License: GPL-3](https://img.shields.io/badge/License-GPL--3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)

Cell type identification remains a key challenge in single-cell RNA-seq and spatial transcriptomics. While many reference-based annotation tools exist, their input and output formats differ greatly, making it cumbersome to run and compare multiple tools.

`AnnotateCells` provides a unified wrapper to run **RCAv2**, **DISCO** and **SingleR** with a single function call, returning consistently formatted outputs that can be directly added to your Seurat object's metadata.


# Installation

Install the required dependencies first:

```{r, eval = FALSE}
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

pacman::p_load(Seurat, SingleR, celldex, tidyverse)

pacman::p_load_gh("satijalab/Azimuth")
pacman::p_load_gh("prabhakarlab/RCAv2")
pacman::p_load_gh("lmw123/DISCOtoolkit")
```

Then install `AnnotateCells`:

```{r, eval = FALSE}
pacman::p_load_gh("adairama/AnnotateCells")
```


# Available tools and panels

<details>
<summary>Click to expand available tools and panels</summary>

```{r, echo = FALSE}
pacman::p_load(tidyverse)
data(allpanels_summary)
knitr::kable(allpanels_summary %>% select(-cell_type) %>% unique() %>% remove_rownames())
```

</details>

For the full list including cell types, see [summary_all_panels.md](https://github.com/adairama/AnnotateCells/blob/main/man/summary_all_panels.md).

Note: `DISCO.all` is also a valid `combo` argument, which annotates against all available DISCO panels listed in the table above.

**References**

- RCAv2:
[Manuscript](https://doi.org/10.1093/nar/gkab632),
[GitHub](https://github.com/prabhakarlab/RCAv2)

- DISCO:
[Manuscript](https://doi.org/10.1093/nar/gkab1020),
[DISCOtoolkit](https://github.com/JinmiaoChenLab/DISCOtoolkit),
[DISCO database](https://www.immunesinglecell.com/),
[Implementation website](https://www.immunesinglecell.com/tool/cellid)

- SingleR:
[Manuscript](https://doi.org/10.1038/s41590-018-0276-y),
[Package](https://doi.org/doi:10.18129/B9.bioc.SingleR),
[celldex package for database](https://doi.org/doi:10.18129/B9.bioc.celldex)


# Notes on annotation

1. The first DISCO annotation run triggers a one-time download of reference data (~175 MB) from the DISCO website, which may take up to 5 minutes. Once downloaded, the data is cached in the `AnnotateCells` package data directory and reused automatically in subsequent runs.

2. DISCO generates predictions at the group level, defined by `Idents()`. These group-level predictions are automatically expanded to the cell level to ensure consistency with other annotation tools in the package.

3. For SingleR, the `combo` argument may be suffixed with `.main`, `.fine`, or `.ont` (e.g. `"SingleR.hpca.main"`, `"SingleR.hpca.fine"`). Entries without a suffix default to `.main` (e.g. `"SingleR.hpca"` → `"SingleR.hpca.main"`).


# Demo dataset

Load the required packages and the `pbmc.demo` dataset:

```{r dataset}
pacman::p_load(tidyverse, Seurat, AnnotateCells)

data(pbmc.demo)
pbmc.demo
```

`pbmc.demo` is a Seurat object containing 2,635 peripheral blood mononuclear cells (PBMC) from a 10X Genomics experiment — the same dataset used in the standard Seurat vignette after QC filtering, with PCA and UMAP embeddings precomputed. See `help(pbmc.demo)` for more details.

Two columns in `meta.data` are relevant here:

- `seurat_annotations`: author-provided cell type labels, used as ground truth for evaluating annotation accuracy
- `RNA_snn_res.0.8`: cluster assignments from the Louvain algorithm at resolution 0.8, yielding 11 clusters (C00–C10) to be annotated

```{r}
pbmc.demo@meta.data %>% head()

# Cross-tabulation of ground truth labels vs clusters
table(pbmc.demo$seurat_annotations, pbmc.demo$RNA_snn_res.0.8)
```


# Running the tool

Pass a vector of `"tool.panel"` string to `AnnotateCells()` to generate cell-level predictions

```{r}
chosen_panels <- c("RCAv2.GlobalPanel_CellTypes",
                   "DISCO.all",
                   "SingleR.hpca.fine")

preds <- AnnotateCells(pbmc.demo, chosen_panels)

head(pred)
```

Then add the result to your Seurat object's metadata:

```{r}
pbmc.demo <- AddMetaData(pbmc.demo, preds)
```


# Aligning predictions from cell-level to cluster-level

The tools above (except DISCO) generate predictions at the individual cell level. Aligning these to cluster groups provides a broader perspective, but a single cluster may contain multiple predicted cell types:

```{r}
table(pbmc.demo$RCAv2.GlobalPanel_CellTypes,
      pbmc.demo$RNA_snn_res.0.8) %>%
  print(zero.print = ".")
```

The `align_prediction_to_cluster()` function visualizes this as a stacked barplot. For each cluster, cell type labels are ordered by frequency (highest at the left) until a cumulative threshold is reached (default: 70%). Labels below the threshold are collapsed into a grey "Misc." bar. Adjust the threshold with `thres` and label size with `text.size` (default: 2).

```{r}
lapply(chosen_panels, function(combo) {
  align_prediction_to_cluster(
    prediction = pbmc.demo@meta.data[, combo],
    cluster    = pbmc.demo$RNA_snn_res.0.8,
    text.size  = 3
  ) + labs(title = combo)
})
```

The function also supports text output formats via the `type` argument (`"split"`, `"long"`, or `"long.all"`):

```{r}
align_prediction_to_cluster(
  prediction = pbmc.demo$RCAv2.GlobalPanel_CellTypes,
  cluster    = pbmc.demo$RNA_snn_res.0.8,
  type       = "split"
)
```
